---
title: "Project proposal: Alaskan Sea Otters"
author: "Autumn Pauly,"
date: "February 2023"
output: github_document
---

```{r load-packages, echo=FALSE, message = FALSE}
library(tidyverse)
library(broom)
```

```{r load-data, echo=FALSE, message=FALSE, warning=FALSE}
base_dir="/Users/pauly/Documents/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/"

seot = read.csv(paste0(base_dir,"seot_morphometricsReproStatus_ak_monson.csv"),h=T)
```

## 1. Introduction
During the 18th and 19th centuries, the commercial harvest of fur extirpated sea otters from their ranges in the North Pacific Ocean. Through reintroduction and protection acts, sea otter populations increased both in distribution and abundance during the 20th century. This resulted in much of their range of Alaska by 2000, excluding their southern ranges. This dataset was taken to assess the sea otter populations along the Alaskan coastlines. 

The dataset included in this project includes morphometric measurements from sea otters that have been collected or captured in Alaska from 1947-2019 by the U.S. Fish and Wildlife Service (USFWS) and U.S. Geological Survey (USGS). Variables within this dataset includes sex, age, reproductive status of female, total body length, tail length, mass, axillary girth, canine diameter, baculum length, paw width, and various other morphological measurements. For females that have pups, the pup’s sex, total length, and mass are included if available to be measured. For harvested animals, data includes fetal sex, length, and mass. Morphometric data from such experimental harvests and captures of sea otters are appropriate for assessing the body size and conditions of this species along a temporal scale. 

## 2. Data
Our data files can be found in the 'data' folder in our project repository on GitHub. 

Data Citation: 
Monson, D. H., Kloecker, K. A. and Esslinger, G. G., 2021, Morphometric and reproductive status data for Sea Otters collected or captured in Alaska: U.S. Geological Survey data release, https://doi.org/10.5066/P9Q5PD3Y.

Referenced Publications(s): 
Tinker, M. T., J. L. Bodkin, L. Bowen, B. E. Ballachey, G. Bentall, A. M. Burdin, H. A. Coletti, G. G. Esslinger, B. B. Hatfield, M. C. Kenner, K. A. Kloecker, B. H. Konar, A. K. Miles, D. H. Monson, M. J. Murray, B. P. Weitzman, and J. A. Estes. 2021. Sea otter population collapse in southwest Alaska: assessing ecological covariates, consequences, and causal factors. Ecological Monographs 91(4):e01472. doi:10.1002/ecm.1472

Laidre, K. L., J. A. Estes, M. T. Tinker, J. L. Bodkin, D. H. Monson, K. Schneider, and M. P. Gillingham. 2006. Patterns of growth and body condition in sea otters from the Aleutian archipelago before and after the recent population decline. Journal of Animal Ecology 75(4):978-989. doi:10.1111/j.1365-2656.2006.01117.x

Monson, D. H. and L. Bowen. 2015. Evaluating the status of individuals and populations: Advantages of multiple approaches and time scales. Pages 121-158 in Larson, S., J. L. Bodkin, and G. R. VanBlaricom (eds) Sea Otter Conservation. Academic Press Boston, MA, USA. doi:10.1016/B978-0-12-801402-8.00006-8

Monson, D. H., J. A. Estes, J. L. Bodkin, and D. B. Siniff. 2000. Life history plasticity and population regulation in sea otters. Oikos 90(3):457-468. doi:10.1034/j.1600-0706.2000.900304.x

## 3. Ethics review
### Limitations in data sources
The data from this project was collected by multiple researchers over several decades (1947-2019). The understanding is that standardized data collection forms were distributed by the USGS to researchers for data entry. Bias, missing values, and uncertainty are always present in data, but in the case of our dataset, there are potentially multiple sources of bias and multiple data collection styles between researchers.

## 4. Data analysis plan
### Data Tidying
The first step in our analysis will have to be to pare down our data files into a more manageable/realistic dataset. This includes filtering and subsetting our data to include the variables that will be needed for our analysis. There have been a variety of inconsistent measurement issues within our dataset, which will have to be corrected through our data tidying efforts. As stated in by the authors, “some experimental harvests (1967-1971) total length was measured as curvilinear length along the back as the otter lay on its stomach and occasionally along the curve of the back while the otters curled in a fetal position due to rigor mortis. In contrast, the length was measured as straight-line length as the otter lay on its back during early studies (1947-1963) and more recent live captures (1986-2019). Therefore, must use corrected “true standard lgth” when comparing lengths between methods” (Monson & Kloecker & Esslinger 2021). This is one such correction that will need to be made. 

### Analysis
The paw width data was collected in an effort to study the estimates of prey size in relation to paw size for sea otter foraging studies. If we are able to find the appropriate datasets, it would be interesting to study the relationship between paw size and prey size. The canine width was measured as a potential metric to sex the individuals. We would like to see if there are any significant differences between male canine lengths and female canine lengths (with respect to the ages of the individuals). The baculum length measurements were taken in an effort to estimate the ages of the male individuals. We would like to perform statistical analysis on if there is a difference in baculum length in males of various ages and if this is an appropriate metric by which to estimate the age of male individuals. We would also like to create a map animation to visualize otter capture locations across the timescale of the dataset.
