---
title: "Project proposal: Alaskan Sea Otters"
author: "Autumn Pauly,"
date: "February 2023"
output: github_document
---

```{r load-packages, echo=FALSE, message = FALSE}
library(tidyverse)
library(broom)
```

```{r load-data, echo=FALSE, message=FALSE, warning=FALSE}
base_dir="/Users/pauly/Documents/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/"

seot = read.csv(paste0(base_dir,"seot_morphometricsReproStatus_ak_monson.csv"),h=T)
```

## 1. Introduction



## 2. Data
Our data files can be found in the 'data' folder in our project repository on GitHub. 

Data Citation: 
Monson, D. H., Kloecker, K. A. and Esslinger, G. G., 2021, Morphometric and reproductive status data for Sea Otters collected or captured in Alaska: U.S. Geological Survey data release, https://doi.org/10.5066/P9Q5PD3Y.

Referenced Publications(s): 
Tinker, M. T., J. L. Bodkin, L. Bowen, B. E. Ballachey, G. Bentall, A. M. Burdin, H. A. Coletti, G. G. Esslinger, B. B. Hatfield, M. C. Kenner, K. A. Kloecker, B. H. Konar, A. K. Miles, D. H. Monson, M. J. Murray, B. P. Weitzman, and J. A. Estes. 2021. Sea otter population collapse in southwest Alaska: assessing ecological covariates, consequences, and causal factors. Ecological Monographs 91(4):e01472. doi:10.1002/ecm.1472

Laidre, K. L., J. A. Estes, M. T. Tinker, J. L. Bodkin, D. H. Monson, K. Schneider, and M. P. Gillingham. 2006. Patterns of growth and body condition in sea otters from the Aleutian archipelago before and after the recent population decline. Journal of Animal Ecology 75(4):978-989. doi:10.1111/j.1365-2656.2006.01117.x

Monson, D. H. and L. Bowen. 2015. Evaluating the status of individuals and populations: Advantages of multiple approaches and time scales. Pages 121-158 in Larson, S., J. L. Bodkin, and G. R. VanBlaricom (eds) Sea Otter Conservation. Academic Press Boston, MA, USA. doi:10.1016/B978-0-12-801402-8.00006-8

Monson, D. H., J. A. Estes, J. L. Bodkin, and D. B. Siniff. 2000. Life history plasticity and population regulation in sea otters. Oikos 90(3):457-468. doi:10.1034/j.1600-0706.2000.900304.x

## 3. Ethics review

## 4. Data analysis plan



