---
title: "Morphometric Analysis"
subtitle: "Of Sea Otters Captured in Alaska (1946-2021)"
author: "Autumn Pauly and Asher Panikian"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    number_sections: true
---
# Initial Setup
## Background Information
[Insert background information here?]

## Loading the Data
First, we'll want to load the appropriate packages and data for the analysis of this dataset.

```{r loading-packages, echo = FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
library(RColorBrewer)
library(stringr)
#install.packages("plotly")
#library(plotly)
#install.packages("gapminder")
#library(gapminder)
```

```{r reading-CSV, echo = FALSE}
#posit cloud
#seot = read.csv("/cloud/project/data/seot.csv", h = T)

#computer
seot = read.csv("~/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/seot.csv", h = T)
```

# Analyzing the Data

### Yearly Observations
Let's look at the yearly observations for the sea otters. As can be shown below, there is a gap in data collection starting from the year 1970 through 1985, which will affect our data ethics. This will need to be accounted for if there is any analysis on temporal differences. 

```{r glimpsing data}
#visualizing the observation per year in a histogram
seot %>% 
  ggplot(aes(x = YEAR, fill = STATE)) +
  geom_histogram(binwidth = 1) +
  theme_grey() +
  scale_fill_manual(values = c("#005a32")) +
  labs(title = "Sea Otter Observations Per Year",
       x = "Year", 
       y = "Count", 
       fill = "")
```
### Tail Length by Sex
#### Creating Plots
We know that sea otter tail lengths are known to be morphometrically different by sex. Let's see if this is true within this dataset. First, let's create a histogram that displays the tail lengths of the male and female otters.

We will want to compare the tail lengths of fully grown adult otters. According to the Alaska Department of Fish and Game, female sea otters become mature around four years of age and male sea otters become mature at age five (Alaska Department of Fish and Game). 

```{r tail-length-sex-histogram}
seot %>% 
  filter(SEX == "M"|SEX == "F") %>% 
  filter(FINAL_AGE > 4) %>% 
  ggplot(mapping = aes(x = mean_tail_lgth, fill = SEX)) + 
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  geom_density() + 
  facet_wrap(~SEX, ncol = 1) + 
  labs(title = "Tail Length (cm) of Adult Otters", 
       subtitle = "by Male and Female",
       x = "Tail Length (cm)", 
       y = "Density", 
       fill = "Sex")
```

#### Statistical Analysis

With the predictor variable (`SEX`) being categorical and the outcome variable (`mean_tail_length`) being quantitative, we will want to use a Welch's Two Sample t-test to determine if there is a difference in tail sizes between sexes.

```{r tail-length-sex-statistics}
#creating data frame with only adult otters
seot_tail <- seot %>% 
  filter(FINAL_AGE > 4)

#Welch's Two Sample t-test
t.test(mean_tail_lgth ~ SEX, data = seot_tail)
```

Report: The tail length of mature male otters (ages 5 and above) is significantly longer than that of mature female otters (Welch Two Sample t-test, t = -2.5123, df = 106.21, p-value = 0.0135).

### BOdy and Tail Length by Sex
#### Creating Plots
Just as with tail lengths, the total lengths of otters have been reported to be different depending on the sex of the otter. Let's see if this is true within this dataset. First, let's create a histogram that displays the total lengths of the male and female otters.

We will again want to filter for the ages of the otters to be representative of the adult population. According to the Alaska Department of Fish and Game, female sea otters become mature around four years of age and male sea otters become mature at age five ([insert citation here]). 

```{r body-length-sex-histogram}
seot %>% 
  filter(SEX == "M"|SEX == "F") %>% 
  filter(FINAL_AGE > 4) %>% 
  ggplot(mapping = aes(x = body_lgth, fill = SEX)) + 
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  geom_density() + 
  facet_wrap(~SEX, ncol = 1) + 
  labs(title = "Total Body Length of Adult Otters", 
       subtitle = "by Male and Female",
       x = "Total Body Length (cm)", 
       y = "Density", 
       fill = "Sex")
```

#### Statistical Analysis

With the predictor variable (`SEX`) being categorical and the outcome variable (`body_lgth`) being quantitative, we will want to use a Welch's Two Sample t-test to determine if there is a difference in total lengths between sexes.

```{r body-length-sex-statistics}
#creating data frame with only adult otters
seot_body <- seot %>% 
  filter(FINAL_AGE > 4)

#Welch's Two Sample t-test
t.test(body_lgth ~ SEX, data = seot_body)
```

Report: The total length of mature male otters (ages 5 and above) is significantly longer than that of mature female otters (Welch Two Sample t-test, t = -14.226, df = 101.57, p-value < 0.0005). Adult male otters have an average body length (from nose to tip of tail) of 104.04 cm while adult female otters have an average body length of 94.85 cm. 

## Age Diversity Among Sex

[add something here about sex diversity in observations?]

```{r age}
seot %>% 
  filter(SEX != "U") %>% 
  ggplot(aes(x=FINAL_AGE, fill = SEX)) + 
   geom_histogram(binwidth = 1) +
   scale_fill_manual(values = c("#addd8e", "#005a32")) +
  theme_grey() +
  xlim(0,25) + 
  facet_grid(~SEX)+ 
  labs(title = "Observations of Ages of Sea Otters", 
       subtitle = "by Male and Female",
       x = "Final Age Assessment", 
       y = "Count", 
       fill = "Sex")
```

## Paw Width by Sex
The paw width data was collected in an effort to study the estimates of prey size in relation to paw size for sea otter foraging studies. If we are able to find the appropriate datasets, it would be interesting to study the relationship between paw size and prey size.

The `PAW` variable is the width of the front paw at the widest point, as measured in millimeters. 

```{r paw-general}
#can I reorder this by region? eastern to western?
seot$REGION <- factor(seot$REGION, levels = c("west_aleutians", "east_aleutians", "alaskan_peninsula", "southeast_alaska", "kodiak", "prince_william_sound"))
```

### Creating Plots

```{r paw-width-age}
#differences in paw width by age


seot %>% 
  filter(PAW > 0) %>% 
  mutate(AGE_CATEGORY = as.character(AGE_CATEGORY)) %>% 
  ggplot(mapping = aes(y = PAW, fill = AGE_CATEGORY)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c("#ffffcc", "#c2e699", "#78c679", "#238443")) +
  facet_wrap(~AGE_CATEGORY, ncol = 4)
```

Further analysis should be done on regional and sex differences. 
```{r paw-width-region}
#differences in paw width by region
seot %>% 
  filter(PAW > 0) %>% 
  filter(REGION != "east_aleutians") %>% 
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6) %>% 
  ggplot(mapping = aes(x = REGION, y = PAW, fill = REGION)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c("#ffffcc", "#d9f0a3", "#addd8e", "#78c679", "#41ab5d", "#238443", "#005a32")) +
  scale_x_discrete(guide = guide_axis(n.dodge = 3)) + 
  labs(title = "Paw Width Measurements of Adult Sea Otters", 
       subtitle = "by Region",
       x = "Region", 
       y = "Paw Width (cm)", 
       fill = "Region")
```

```{r adult-paw-width-sex}
seot %>% 
  filter(PAW > 0) %>%
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6) %>%  
  ggplot(mapping = aes(x = PAW, fill = SEX)) +
  geom_histogram() + 
  facet_wrap(~SEX, ncol = 1) + 
   scale_fill_manual(values = c("#addd8e", "#005a32")) + 
  labs(title = "Paw Width in Sea Otters", 
       subtitle = "by Male and Female",
       x = "Paw Width (mm)", 
       y = "Count", 
       fill = "Sex")
```
#### Statistical Analysis 

```{r ANOVA-paw-size}
seot_paw <- seot %>% 
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6)

pawaov <- aov(PAW ~ SEX, data = seot)
summary(pawaov)
TukeyHSD(pawaov)
```

REPORT: The paw size of female otters was signficantly different than that of female otters [insert test statistic here]. 

Is this the test to use? 


## Canine and Sexing
The canine width was measured as a potential metric to sex the individuals. We would like to see if there are any significant differences between male canine lengths and female canine lengths (with respect to the ages of the individuals).

The canine measurement is labeled under the `CAN_DIA` column, where a measurement of the diameter of canine tooth (mm) was taken at the gum line, which is the widest dimension. This canine width was measured with a caliper and recorded to the nearest 0.1 mm. 

### Creating Plots
Let's create a plot to visualize the distribution of the data. This will help us determine which type of statistical test will be the most appropriate to use. First, we have to filter the `seot` dataset so that we will be able to visualize the `CAN_DIA` and `SEX` variables that we're interested in. 

```{r glimpsing-canine-measurement-by-sex}
seot_sex <- seot %>% 
  mutate(CAN_DIA = as.numeric(CAN_DIA)) %>% 
  filter(SEX != "U") %>% 
  filter(CAN_DIA > 0) %>% 
  filter(FINAL_AGE > 0)

#boxplot
ggplot(data = seot_sex, mapping = aes(x = SEX, y = CAN_DIA, fill = SEX)) + 
  geom_boxplot() + 
  scale_fill_manual(values = c("#addd8e", "#005a32")) +
  ylim(2,12) + 
  labs(title = "Canine Diameter by Sex",
       x = "Sex",
       y = "Diameter (mm)", 
       color = "Sex")
```

```{r scatterplot-canine}
#scatterplot
ggplot(data = seot_sex, mapping = aes(x = FINAL_AGE, y = CAN_DIA, color = CAN_DIA)) + 
  geom_jitter() + 
  ylim(5,12) + 
  geom_smooth(color = "black") +
  facet_grid(~SEX) +
  scale_color_viridis_c() + 
  labs(title = "Canine Diameter by Sex",
       x = "Sex",
       y = "Diameter (mm)", 
       color = "Sex")

```

```{r canine-histogram}
#histogram
ggplot(data = seot_sex, mapping = aes(x = CAN_DIA, fill = SEX)) +
  geom_histogram() + 
  facet_grid(~SEX) + 
  scale_fill_manual(values = c("#addd8e", "#005a32")) +
  xlim(2,13) + 
  labs( title = "Canine Diameter of Sea Otters", 
        subtitle = "by Sex", 
        x = "Canine Diameter (mm)", 
        y = "Count", 
        fill = "Sex")
```

### Statistical Analysis

To truly determine if the measurement of the canine diameter at the gum line is an appropriate metric by which to sex a sea otter, statistical testing should be performed. As was seen above, the distribution of both sexes appears to conform to a normal distribution. This suggests that we can use a test that follows a two-sample t-test with equal variance. 

```{r stats-canine-ttest}
t.test(CAN_DIA ~ SEX, data = seot_sex, var.equal = TRUE)
```

The diameter of the canine tooth for female sea otters is significantly smaller than that of the male otters (Two Sample t-test; t = -19.164, df = 1132, p-value < 2.2e-16). This would mean that this would be an appropriate metric by which to determine the sex of an unknown sea otter. 

#### Modeling Canine Diameter
```{r install-tidymodels}
# install.packages("tidymodels")
# library(tidymodels)
```

For each additional year, the male otter's baculum is expected to grow by 0.22 cm. For otters that are 0 (aka newborn), the baculum length is expected to be 14.33 cm. 
```{r model-canine-vs-sex}
# #first, we need to filter out the na values (that being -9.0)
# seot_canine <- seot %>% 
#   filter(CAN_DIA > 0)
# 
# #now, let's create the linear model comparing final age to baculum length
# seot_model_canine <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(CAN_DIA ~ SEX, data = seot_canine)
# 
# #let's tidy this model
# seot_model_canine %>% tidy()
```

```{r graphical-diagnostics-canine}
# #augmenting the baculum data
# seot_model_canine_augment <- augment(seot_model_canine$fit)
# 
# ggplot(seot_model_canine_augment, mapping = aes(x = .fitted, y = .resid, color = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   scale_color_viridis_c() +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted Age", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_model_canine)$adj.r.squared
```
## Baculum Length and Aging

The baculum length was measured on males a way to estimate age-class of the animal. We would like to perform statistical analysis on if there is a difference in baculum length in males of various ages and if this is an appropriate metric by which to estimate the age of male individuals.

The baculum measurement is found under the `BACULA_LGTH` column with the unit of measurement being centimeters. The baculum length was measure using a tape measure and feeling the ends of the bone and recorded to the nearest 0.5 cm.

### Creating Plots
Let's create a plot to visualize the distribution of the data. This will help us determine which type of statistical test will be the most appropriate to use. First, we have to filter the `seot` dataset so that we will be able to visualize the `BACULA_LGTH` and `FINAL_AGE` variables that we're interested in. 

```{r baculum-length-visualization}
seot_baculum <- seot %>% 
  filter(SEX == "M") %>% 
  mutate(FINAL_AGE = as.numeric(FINAL_AGE)) %>% 
  filter(FINAL_AGE > 0) %>% 
  mutate(BACULA_LGTH = as.numeric(BACULA_LGTH)) %>% 
  filter(BACULA_LGTH > 0)

ggplot(data = seot_baculum, mapping = aes(x = FINAL_AGE, y = BACULA_LGTH, color = FINAL_AGE)) +
  geom_jitter() + 
  scale_color_viridis_c() +
  geom_smooth(color = "black") +
  labs(title = "Baculum Length by Age", 
       x = "Final Age", 
       y = "Bacula Length (cm)", 
       color = "Age")
```

### Statistical Analysis

The statistics that we find with the Tukey HSD test that we performed is a little convoluted, so let's tidy this up and group ages into categories and try again. 

```{r tukeyhsd-baculumlength-finalage, message=FALSE}
#create dataframe that includes the variables we want
seot_baculum <- seot_baculum %>% 
  mutate(FINAL_AGE = as.factor(FINAL_AGE)) %>% 
  select(BACULA_LGTH, FINAL_AGE) 

#creating the ANOVA test
aovbaculum <- aov(BACULA_LGTH~FINAL_AGE, data = seot_baculum)
summary(aovbaculum)

#Tukey HSD Test
TukeyHSD(aovbaculum)
```

Now we can more clearly see how the baculum length relates to ages. 

```{r tukeyhsd-baculumlength-agecategory}
#create dataframe that includes the variables we want
seot_baculum_2 <- seot %>% 
  filter(SEX == "M") %>% 
  mutate(AGE_CATEGORY = as.numeric(AGE_CATEGORY)) %>% 
  filter(AGE_CATEGORY > 0) %>% 
  mutate(BACULA_LGTH = as.numeric(BACULA_LGTH)) %>% 
  filter(BACULA_LGTH > 0)

seot_baculum_2 <- seot_baculum_2 %>% 
  mutate(AGE_CATEGORY = as.factor(AGE_CATEGORY)) %>% 
  select(BACULA_LGTH, AGE_CATEGORY) 

#creating the ANOVA test
aovbaculum2 <- aov(BACULA_LGTH~AGE_CATEGORY, data = seot_baculum_2)
summary(aovbaculum2)

#Tukey HSD Test
TukeyHSD(aovbaculum2)
```

REPORT: The baculum length of individuals who are ages 1 - 7 are significantly different than one another (one-way ANOVA, F_2,457, = 112, P < 0.005). In a Tukey HSD post-hoc test, the baculum length of individuals who were from ages 7 to 13 did not differ significantly (refer to Tukey multiple comparisons of means). This would suggest that you could use the baculum length to estimate the age of individuals who were 7 years or younger, but not for individuals who are over 7 years old. 

### Creating a Model of Baculum Length by Age

For each additional year, the male otter's baculum is expected to grow by 0.22 cm. For otters that are 0 (aka newborn), the baculum length is expected to be 14.33 cm. 
```{r model-baculum-vs-age}
# #first, we need to filter out the na values (that being -9.0)
# seot_bacula <- seot %>% 
#   filter(BACULA_LGTH > 0)
# 
# #now, let's create the linear model comparing final age to baculum length
# seot_model_baculum <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(BACULA_LGTH ~ FINAL_AGE, data = seot_bacula)
# 
# #let's tidy this model
# seot_model_baculum %>% tidy()
```

```{r graphical-diagnostics-baculum}
# #augmenting the baculum data
# seot_model_baculum_augment <- augment(seot_model_baculum$fit)
# 
# ggplot(seot_model_baculum_augment, mapping = aes(x = .fitted, y = .resid, color = .fitted)) +
#   geom_jitter(alpha = 0.75) +
#   scale_color_viridis_c() +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted Age", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_model_baculum)$adj.r.squared
```

## Weight by Tail Length
[insert background information here]

### Creating Plots
```{r weight-tail-length, warning=FALSE}
#creating a plot that visualizes the relationship between weight and tail length
ggplot(data = seot, mapping = aes(x = WEIGHT, y = mean_tail_lgth, color = WEIGHT)) + 
  geom_point(alpha = 0.75) +
  scale_color_viridis_c() +
  ylim(15, 40) + 
  xlim(0, 45) + 
  geom_smooth(color = "black") + 
  labs(x = "Weight (kg)", 
       y = "Mean Tail Length (cm)", 
       color = "Weight")
```

### Statistical Analysis

With the tail length and the weight of the otter being both quantitative variables, we should use a simple regression statistical test to determine if there is a statistical significance. 

```{r linear-model-tail-weight}
#creating the linear model
weight.tail.lm <- lm(mean_tail_lgth ~ WEIGHT, data = seot)

summary(weight.tail.lm)
```

REPORT: There is a significant relationship (p < 0.001) between tail length and weight of adult otters (R2 = 2.14 ± 0.014), with a 0.23-mm increase in reported tail length for every year increase in the otter's age.

The regression equation is: 
tail length = 24.41 + 0.23(weight) ± 0.014. 

## Recapture Data 

```{r recaptured-otters, include=FALSE}
# seot_recapture <- seot %>% 
#   select(YEAR, OTTER.NO, FINAL_AGE, WEIGHT, SEX) %>%
#   mutate(OTTER.NO = as.factor(OTTER.NO))
```

```{r recapture-data-again, include=FALSE}
# seot_recapture %>% 
#   filter(OTTER.NO == "SO-01-28") %>% 
#   select(YEAR, OTTER.NO, FINAL_AGE, SEX, WEIGHT)
# 
# seot_recapture %>% 
#   filter(OTTER.NO == "SO-02-07") %>% 
#   select(YEAR, OTTER.NO, FINAL_AGE, SEX, WEIGHT)
# 
# seot_recapture %>% 
#   filter(FINAL_AGE > 0) %>% 
#   select(YEAR, OTTER.NO, FINAL_AGE, SEX)
```
