---
title: "Data Tidying"
author: "Asher Panikian and Autumn Pauly"
output: github_document
date:
---

# Loading Data
```{r loading-packages}
library(tidyverse)
```

```{r loading-dataset}
#base_dir="/Users/pauly/Documents/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/"
seot = read.csv("/cloud/project/data/seot_morphometricsReproStatus_ak_monson.csv", h = T)

#for Autumn's computer directory
#seot = read.csv("~/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/seot_morphometricsReproStatus_ak_monson.csv", h = T)
```

# Glimpsing Data
```{r glimpsing data}
glimpse(seot)
```

A lot of the variables that are included in this dataset will not be used for analysis - a variety of these variables are used to determine the value of another variable (i.e. how `LGTH_1` is used to determine ``body_lgth`) and as such we will not need them in our analysis. We need to then select out which variables we will want to keep. 

```{r selecting-variables}
#selecting the variables that we want to keep
seot <- seot %>% 
  select(OTTER.NO, recap, DATE, YEAR, LOCATION, STATE, REGION, AREA, LAT, LONG, SEX, WEIGHT, true_standard_lgth, mean_tail_lgth, body_lgth, mean_girth, PAW, W_PUP, PUP_NUMBER, PUP_SEX, PUP_WGHT, PUP_LGTH, PUP_CURVLGTH, FETUS_PRES, FETUS_NUM, FETUS_SEX, FETUS_WT, FETUS_LTH, FE_REP_CON, FE_REP_STA, PREGNANCY_STATUS, CAN_DIA, AGE_CATEGORY, FINAL_AGE, BACULA_LGTH, Cause_of_death_Capture_method)
```


# Spatial Naming
### Changing Regional Names
A number of the regional names within this dataset are shortened, as you can see below. 

```{r unique-regional-names}
unique(seot$REGION)
```

We would like to rename these values so that the names are more intuitive yet still concise. This will help us with our data analysis and understanding, as neither of the project members are from Alaska nor are we familiar with what the areas the shortened names are referring to. 

```{r changing-region-names}
#changing region variable to factor
seot <- seot %>% 
  mutate(REGION = as.factor(REGION))

#changing regional names
seot <- seot %>% 
  mutate(REGION = recode(REGION, 
                         "AK peninsula" = "alaskan_peninsula", 
                         "PWS" = "prince_william_sound", 
                         "SE" = "southeast_alaska", 
                         "east aleutians" = "east_aleutians", 
                         "lower cook inlet" = "lower_cook_inlet", 
                         "west aleutians" = "west_aleutians"))
```

As we look at the number of observations per region, we can find that there is a large difference between the `west_aleutians` and the `east_aleutians`. This may prove to be an issue with spatial analysis, but further investigation into the locality of the otter capture will determine if any action is needed on the drastic difference in observations per region. 

```{r counting-region}
seot %>% 
  count(REGION) %>% 
  arrange(desc(n))
```

### Changing Area Names

As was with the regional names, the area names are also shortened for ease of data entry. We will be changing these names to increase our subjective, intuitive analysis of this dataset. 

```{r changing-area-names}
#changing area variable to factor
seot <- seot %>% 
  mutate(AREA = as.factor(AREA))

#changing area names
seot <- seot %>% 
  mutate(AREA = recode(AREA, 
                         "CSE" = "central_southeast_alaska" , 
                          "EAP" = "eastern_alaskan_peninsula", 
                         "EPWS" = "eastern_prince_william_sound", 
                          "NSE" = "northern_southeast_alaska", 
                          "SSE" = "southern_southeast_alaska", 
                          "WAP" = "western_alaskan_peninsula", 
                          "WPWS" = "western_prince_william_sound", 
                          "fox islands" =  "fox_islands", 
                          "near islands" = "near_islands", 
                          "afognak" = "afognak_island", 
                           "andreanofs" = "andreanof_islands", 
                         "delarofs" = "delarof_islands" , 
                         "kachemak" = "kachemak_bay", 
                          "kodiak" = "kodiak_island" , 
                          "rats" = "rat_islands" , 
                          "shuyak" = "shuyak_island_kodiak" ))
```

Much like the regional observations, there is a large difference between the number of observations per specific location. This will also need to be assessed to see if any further action will be needed for spatial analysis.

```{r counting-area}
seot %>% 
  count(AREA) %>% 
  arrange(desc(n))
```

# Cleaning up NA values

As shown in the codebook, the NA values found within this dataset are usually a negative integer. 

```{r unique-negative-values}
range(seot$WEIGHT)
range(seot$true_standard_lgth)
range(seot$mean_tail_lgth)
range(seot$body_lgth)
range(seot$mean_girth)
range(seot$BACULA_LGTH)

#and it goes on...
```

Let's change this so that the values are truly NA.

```{r making-na-values}
seot <- seot %>% 
#weight of otter
  mutate(WEIGHT = if_else(WEIGHT < 0, NA, WEIGHT)) %>% 
#true standard length of total body
  mutate(true_standard_lgth = if_else(true_standard_lgth < 0, NA, true_standard_lgth)) %>%
#body length (true standard length - tail length)
  mutate(body_lgth = if_else(body_lgth < 0, NA, body_lgth)) %>%
#mean girth of body
  mutate(mean_girth = if_else(mean_girth < 0, NA, mean_girth)) %>%
#paw length  
  mutate(PAW = if_else(PAW < 0, NA, PAW)) %>%
#canine diameter
  mutate(CAN_DIA = if_else(CAN_DIA < 0, NA, CAN_DIA)) %>%
#final age
  mutate(FINAL_AGE = if_else(FINAL_AGE < 0, NA, FINAL_AGE)) %>%
#bacula length
  mutate(BACULA_LGTH = if_else(BACULA_LGTH < 0, NA, BACULA_LGTH)) %>% 
#mean tail length
  mutate(mean_tail_lgth = if_else(mean_tail_lgth < 0, NA, mean_tail_lgth)) %>% 
#sex of otters
  mutate(SEX = if_else(SEX == "U", NA, SEX)) %>%
#age category
  mutate(AGE_CATEGORY = if_else(AGE_CATEGORY < 0, NA, AGE_CATEGORY))
```

Problem solved!

## Dates to Date Format

```{r date-to-date}
seot <- seot %>% 
  mutate(DATE = as.Date(DATE))
```


# Writing CSV

Now, we should write this new tidied dataset as a CSV so that it can be read across all the .Rmd files. 

```{r CSV-writing}
write_csv(seot, "/cloud/project/data/seot.csv")
```
