---
title: "Oil Spill Analysis"
author: "Autumn Pauly and Asher Panikian"
output: github_document
---

# Background Information
These data within the USFWS and USGS dataset was collected primarily during experimental harvests primarily conducted in the 1960's and 1970's, or during sea otter captures related to USFWS and USGS sea otter sampling and tagging studies conducted from the late 1980's up to and including 2019, or recovered dead during the 1989 Exxon Valdez Oil spill (EVOS).

We were interested to see what types of otters were found dead after the 1989 Exxon Valdez Oil spill (EVOS) and if there were any commonalities between these deceased individuals. 

# Loading the Data
First, we'll want to load the appropriate packages and data for the analysis of this dataset.

```{r loading-packages, echo = FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
library(RColorBrewer)
library(stringr)
#install.packages("plotly")
#library(plotly)
#install.packages("gapminder")
#library(gapminder)
```

```{r reading-CSV, echo = FALSE}
#posit cloud
seot = read.csv("/cloud/project/data/seot.csv", h = T)

#computer
#seot = read.csv("~/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/seot.csv", h = T)
```

# Analyzing the Data

## Otters Found Dead
Let's look and see which otters were found dead and how many we have. 

In this dataset, the value `1` is defined as "Collected via harvest", `0` is defined as "Typical live capture", `4` is defined as "Measured after some time in captivity", and `5` is defined as "Captured off a beach". We will be looking for the varibles where they were captured via harvest. 

### Otter's Weight before and after the Exxon Oil Spill
#### Creating Plots

```{r glimpsing-dead-otters-weight}
#sea otters after the oil spill
seot <- seot %>% 
  mutate(oil_spill = case_when(YEAR < 1989 ~ "before_oil_spill", 
                               YEAR > 1989 ~ "after_oil_spill"))

#adult weight
seot %>% 
  filter(FINAL_AGE > 4) %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill)) + 
  geom_boxplot() +
  labs(title = "Adult Sea Otter Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")

#otters aged 4 weight
seot %>% 
  filter(FINAL_AGE == "4") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill)) + 
  geom_boxplot() +
  labs(title = "Sea Otter Aged 4 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")

#otters aged 3 weight
seot %>% 
  filter(FINAL_AGE == "3") %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill)) + 
  geom_boxplot() +
  labs(title = "Sea Otter Aged 3 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")

#otters aged 2 weight
seot %>% 
  filter(FINAL_AGE == "2") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill)) + 
  geom_boxplot() +
  labs(title = "Sea Otter Aged 2 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")

#otters aged 1 weight
seot %>% 
  filter(FINAL_AGE == "1") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill)) + 
  geom_boxplot() +
  labs(title = "Sea Otter Aged 1 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```

#### Check the R-Squared

```{r}

```

#### Statistical Analysis

```{r}

```

### Otter's morphometric measurements before and after the Exxon Oil Spill
#### Creating Plots

```{r}

```

#### Check the R-Squared

```{r}

```

#### Statistical Analysis

```{r}

```






















