---
title: "Oil Spill Analysis"
author: "Autumn Pauly and Asher Panikian"
output: 
  html_document:
    toc: true
    number_sections: true
---

# Background Information
These data within the USFWS and USGS dataset was collected primarily during experimental harvests primarily conducted in the 1960's and 1970's, or during sea otter captures related to USFWS and USGS sea otter sampling and tagging studies conducted from the late 1980's up to and including 2019, or recovered dead during the 1989 Exxon Valdez Oil spill (EVOS).

We were interested to see if there were any differences between the otters that were captured before and after the oil spill.

# Loading the Data
First, we'll want to load the appropriate packages and data for the analysis of this dataset.

```{r loading-packages, echo = FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
library(RColorBrewer)
library(stringr)
#install.packages("plotly")
#library(plotly)
#install.packages("gapminder")
#library(gapminder)
```

```{r reading-CSV, warning=FALSE}
#posit cloud
seot = read.csv("/cloud/project/data/seot.csv", h = T)

#computer
#seot = read.csv("~/GitHub/project-ds-2-autumn/project-ds-2-autumn/data/seot.csv", h = T)
```

First, we'll want to create a new variable called `oil_spill` where the dataset will be split into `before_oil_spill` and `after_oil_spill`. This variable will also be factored so that the years will be in chronological order. 

```{r factoring-oil-spill-status, warning=FALSE}
#sea otters after the oil spill
seot <- seot %>% 
  mutate(oil_spill = case_when(YEAR < 1989 ~ "before_oil_spill", 
                               YEAR > 1989 ~ "after_oil_spill"))

seot$oil_spill <- factor(seot$oil_spill, levels = c("before_oil_spill", "after_oil_spill"))
```












# Analyzing the Data
### General Otter's Weight before and after the Exxon Oil Spill
#### Creating Plots

This graph visually represents the difference in adult sea otter weight before and after the Exxon Oil Spill in 1989 in Prince William's Sound. 

```{r adult-weight, warning=FALSE}
#adult weight
seot %>% 
  filter(FINAL_AGE > 4) %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  labs(title = "Adult Sea Otter Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "Status", 
       y = "Weight (lbs)",
       fill = "Status")

```

This graph visually represents the difference in sea otters aged 4 years old weight before and after the Exxon Oil Spill in 1989 in Prince William's Sound. 

```{r otter-age-4-weight, warning=FALSE}
#otters aged 4 weight
seot %>% 
  filter(FINAL_AGE == "4") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  labs(title = "Sea Otter Aged 4 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "Status", 
       y = "Weight (lbs)",
       fill = "Status")
```

This graph visually represents the difference in sea otters aged 3 years old weight before and after the Exxon Oil Spill in 1989 in Prince William's Sound. 

```{r otter-age-3-weight, warning=FALSE}
#otters aged 3 weight
seot %>% 
  filter(FINAL_AGE == "3") %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  labs(title = "Sea Otter Aged 3 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "Status", 
       y = "Weight (lbs)",
       fill = "Status")
```

This graph visually represents the difference in sea otters aged 2 years old weight before and after the Exxon Oil Spill in 1989 in Prince William's Sound. 

```{r otter-age-2-weight, warning=FALSE}
#otters aged 2 weight
seot %>% 
  filter(FINAL_AGE == "2") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  labs(title = "Sea Otter Aged 2 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "Status", 
       y = "Weight (lbs)",
       fill = "Status")
```

This graph visually represents the difference in sea otters aged 1 years old weight before and after the Exxon Oil Spill in 1989 in Prince William's Sound. 

```{r otter-age-1-weight, warning=FALSE}
#otters aged 1 weight
seot %>% 
  filter(FINAL_AGE == "1") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#d9f0a3", "#005a32")) +
  labs(title = "Sea Otter Aged 1 Weight", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "Status", 
       y = "Weight (lbs)",
       fill = "Status")
```

#### Check the R-Squared

```{r r-squared-weight-oil}
# #now, let's create the linear model comparing weight to year
# seot_model_oilweight <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(WEIGHT ~ oil_spill, data = seot)
# 
# #let's tidy this model
# seot_model_oilweight %>% tidy()
# 
# #augmenting the weight data
# seot_model_oilweight_augment <- augment(seot_model_oilweight$fit)
# 
# ggplot(seot_model_oilweight_augment, mapping = aes(x = .fitted, y = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted weight", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_model_oilweight)$adj.r.squared
```

The r-squared value is `0.03906333`, which means that the variable `oil_spill` accounts for 39% of the variation in the dependent variable, `WEIGHT`. 

#### Statistical Analysis

```{r stats-weight-oil}
#Welch's Two Sample t-test
t_test_weight <- t.test(WEIGHT ~ oil_spill, data = seot)
t_test_weight

#two-way ANOVA
two_anova_weight_oil <- aov(WEIGHT ~ oil_spill + FINAL_AGE + oil_spill:FINAL_AGE, data = seot)
summary(two_anova_weight_oil)

#interaction plot of data
interaction.plot(x.factor=seot$FINAL_AGE, trace.factor = seot$oil_spill, response = seot$WEIGHT, type = "o",pch=16)
```

The two-way ANOVA suggests that there is a statistical significance with the interaction between the two variables `FINAL_AGE` and `oil_spill` for the dependent variable `WEIGHT`, we will have to perform any analysis of simple events.

```{r simple-events-analysis-weight}
TukeyHSD(two_anova_weight_oil, "WEIGHT")
```














### Specific Otter's Weight before and after the Exxon Oil Spill
We have already looked to see if the general weight of the captured sea otters were significantly different before and after the Exxon oil spill, but now we'd like to see if there were any regional differences between the weights. 

#### Creating Plots
As shown below, the West Aleutians are one of the only sites where there were otters captured both before and after the oil spill. As such, we will be analyzing this region of Alaska. 

```{r weight-specific}
seot %>%
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6) %>% 
  ggplot(mapping = aes(x = YEAR)) + 
  geom_histogram() + 
  facet_wrap(~REGION) + 
  labs(title = "Otters Observed Through the Years", 
       subtitle = "by Region", 
       x = "Year", 
       y = "Count")
```

```{r west_aleutians-filter}
seot_west_aleutians <- seot %>% 
  filter(REGION == "west_aleutians")
```

```{r west-aleutian-weight-by-region}
seot_west_aleutians %>%
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6) %>% 
  ggplot(mapping = aes(y = WEIGHT, x = YEAR, color = oil_spill)) + 
  geom_jitter() +
  scale_color_manual(values = c("#78c679", "#005a32")) +
  geom_smooth(color = "black") +
  labs(title = "Otters Observed Through the Years", 
       subtitle = "in the West Aleutians", 
       x = "Year", 
       y = "Count", 
       color = "Status")
```

```{r west-aleutian-weight-by-region-boxplot}
seot_west_aleutians %>%
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6) %>% 
  ggplot(mapping = aes(y = WEIGHT, x = oil_spill, fill = oil_spill)) + 
  scale_fill_manual(values = c("#78c679", "#005a32")) +
  geom_boxplot() +
  labs(title = "Otters Observed Through the Years", 
       subtitle = "in the West Aleutians", 
       x = "Status", 
       y = "Count", 
       fill = "Status")
```

#### Check the R-Squared

```{r r-squared-weight-oil-west-aluetian}
# #now, let's create the linear model comparing weight to year
# seot_wa_model_oilweight <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(WEIGHT ~ oil_spill, data = seot_west_aleutians)
# 
# #let's tidy this model
# seot_wa_model_oilweight %>% tidy()
# 
# #augmenting the weight data
# seot_wa_model_oilweight_augment <- augment(seot_wa_model_oilweight$fit)
# 
# ggplot(seot_wa_model_oilweight_augment, mapping = aes(x = .fitted, y = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted weight", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_wa_model_oilweight)$adj.r.squared
```
REPORT: The r-squared value is `0.01960116`, which means that the variable `oil_spill` accounts for 1.96% of the variation in the dependent variable, `WEIGHT`. 

```{r new-model-wa}
# #now, let's create the linear model comparing weight to year
# seot_new_wa_model_oilweight <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(WEIGHT ~ oil_spill, data = seot_west_aleutians)
# 
# #augmenting the weight data
# seot_new_wa_model_oilweight_augment <- augment(seot_new_wa_model_oilweight$fit)
# 
# ggplot(seot_new_wa_model_oilweight_augment, mapping = aes(x = .fitted, y = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted weight", y = "Residuals")
# 
# #assessing the new r-squared value
# glance(seot_new_wa_model_oilweight)$adj.r.squared
```

REPORT: The r-squared value is `0.4123082`, which means that the variable `oil_spill` along with the variable `FINAL_AGE` accounts for 41.23% of the variation in the dependent variable, `WEIGHT`. 

Now, we would like to assess the r-squared value of the variable `oil_spill` on the adult sea otter populations in the Western Aleutians.

```{r new-model-adults-wa}
# seot_west_aleutians_adult <- seot_west_aleutians %>% 
#   filter(FINAL_AGE > 4|AGE_CATEGORY > 6)
# 
# #now, let's create the linear model comparing weight to year
# seot_adult_wa_model_oilweight <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(WEIGHT ~ oil_spill, data = seot_west_aleutians_adult)
# 
# #let's tidy this model
# seot_adult_wa_model_oilweight %>% tidy()
# 
# #augmenting the weight data
# seot_adult_wa_model_oilweight_augment <- augment(seot_adult_wa_model_oilweight$fit)
# 
# ggplot(seot_adult_wa_model_oilweight_augment, mapping = aes(x = .fitted, y = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted weight", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_adult_wa_model_oilweight)$adj.r.squared
```

#### Statistical Analysis

```{r stats-weight-oil-west-aluetian}
#Welch's Two Sample t-test
t_test_wa_weight <- t.test(WEIGHT ~ oil_spill, data = seot_west_aleutians)
t_test_wa_weight
```

REPORT: There was a significant difference before and after the oil spill of the weight of the adult sea otter population in the Western Aleutians, where they were significantly heavier before the oil spill (Welch Two Sample t-test; t = 7.4412, df = 399.18, p-value < 0.005). An assessment of the r-squared value suggests that the variable `oil_spill` accounts for 1.96% of the variation in the dependent variable, `WEIGHT` in the western Aluetians.







### Otter's Body Length before and after the Exxon Oil Spill
#### Creating Plots

This is a graph for adult otter body lengths before and after the Exxon Oil Spill. 

```{r body_length-oil, echo = FALSE}
#adult body length
seot_west_aleutians %>% 
  filter(FINAL_AGE > 4|AGE_CATEGORY > 6) %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>%
  ggplot(mapping = aes(y = true_standard_lgth, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
    scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Adult Sea Otter Body Length", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```

This is a graph for sea otters aged 4 body lengths before and after the Exxon Oil Spill. 

```{r age-4-otter-length, echo = FALSE}
#otters aged 4 body length
seot_west_aleutians %>% 
  filter(FINAL_AGE == "4") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = true_standard_lgth, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
    scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Sea Otter Aged 4 Body Length", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```

This is a graph for sea otters aged 3 body lengths before and after the Exxon Oil Spill. 

```{r age-3-otter-length, echo = FALSE}
#otters aged 3 body length
seot_west_aleutians %>% 
  filter(FINAL_AGE == "3") %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = true_standard_lgth, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
    scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Sea Otter Aged 3 Body Length", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```

This is a graph for sea otters aged 2 body lengths before and after the Exxon Oil Spill. 

```{r age-2-otter-length, echo = FALSE}
#otters aged 2 body length
seot_west_aleutians %>% 
  filter(FINAL_AGE == "2") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>%
  ggplot(mapping = aes(y = true_standard_lgth, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
    scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Sea Otter Aged 2 Body Length", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```

This is a graph for sea otters aged 1 body lengths before and after the Exxon Oil Spill. 

```{r age-1-otter-length, echo = FALSE}
#otters aged 1 body length
seot_west_aleutians %>% 
  filter(FINAL_AGE == "1") %>%
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>%  
  ggplot(mapping = aes(y = true_standard_lgth, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
    scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Sea Otter Aged 1 Body Length", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```


#### Check the R-Squared

```{r stats-length-oilspill}
# #now, let's create the linear model comparing weight to year
# seot_model_oillength <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(true_standard_lgth ~ oil_spill, data = seot_west_aleutians)
# 
# #let's tidy this model
# seot_model_oillength %>% tidy()
# 
# #augmenting the weight data
# seot_model_oillength_augment <- augment(seot_model_oillength$fit)
# 
# ggplot(seot_model_oillength_augment, mapping = aes(x = .fitted, y = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted weight", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_model_oillength)$adj.r.squared
```

#### Statistical Analysis

We are performing a statistical analysis on the body length of adult otters before and after the Exxon Oil Spill. This is an analysis of two variables (one qualitative, one quantitative), where there are two groups being compared against parametric, independent samples. This suggests that we should use a Welch Two Sample t-test for statistical analysis.

```{r oil-weight-statstest}
#Welch's Two Sample t-test
t_test_length <- t.test(true_standard_lgth ~ oil_spill, data = seot_west_aleutians)
t_test_length
```

REPORT: There was no significant impact of the oil spill on the body length of the sea otter population (Welch Two Sample t-test; t = -1.0462, df = 350.38, p-value = 0.2962). There was no significant impact of the oil spill on the body length of the sea otter population (Two-way ANOVA; F_1,3375, f = 0.207, p=0.65). There was a significant impact on the age of the otter on the body length (Two-way ANOVA; F_1,3375, f = 2297.583, p < 0.0005).















### Adult Otter's Baculum Length before and after the Exxon Oil Spill
#### Creating Plots

```{r bacula_length-oil}
#adult body length
seot_west_aleutians %>% 
  filter(FINAL_AGE > 4) %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>%
  filter(REGION == "west_aleutians") %>% 
  ggplot(mapping = aes(y = BACULA_LGTH, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Adult Sea Otter Bacula Length", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")
```

#### Check the R-Squared

```{r stats-weight-oilspill}
#now, let's create the linear model comparing weight to year
seot_model_baclength <- linear_reg() %>%
  set_engine("lm") %>%
  fit(BACULA_LGTH ~ oil_spill, data = seot_west_aleutians)

#let's tidy this model
seot_model_baclength %>% tidy()

#augmenting the weight data
seot_model_baclength_augment <- augment(seot_model_baclength$fit)

ggplot(seot_model_baclength_augment, mapping = aes(x = .fitted, y = .resid)) +
  geom_jitter(alpha = 0.75) +
  geom_smooth(color = "black") +
  labs(x = "Predicted weight", y = "Residuals")

#assessing the r-squared value
glance(seot_model_baclength)$adj.r.squared
```

#### Statistical Analysis

We are performing a statistical analysis on the body length of adult otters before and after the Exxon Oil Spill. This is an analysis of two variables (one qualitative, one quantitative), where there are two groups being compared against parametric, independent samples. This suggests that we should use a Welch Two Sample t-test for statistical analysis.

```{r specific-oil-length-statstest}
#Welch's Two Sample t-test
t_test_length <- t.test(BACULA_LGTH ~ oil_spill, data = seot_west_aleutians)
t_test_length

#two-way ANOVA
two_anova_lgth_bac <- aov(BACULA_LGTH ~ oil_spill + FINAL_AGE + oil_spill:FINAL_AGE, data = seot_west_aleutians)
summary(two_anova_lgth_bac)
```

With there being no statistical significance with the interaction between the two variables `FINAL_AGE` and `oil_spill` for the dependent variable `BACULA_LGTH`, we don't have to perform any analysis of simple events. 

REPORT: There was a significant difference in baculum sizes in adult male otters before and after the Exxon Mobil oil spill in 1989 (Welch Two Sample t-test; t = 4.6688, df = 183.77, p-value = 0.005). The average baculum length for adult male otters before the oil spill was 14.02 mm and 15.85 mm after the oil spill. 

### Temporal Time Scale
We want to see if there was a noticale change in baculum length over time. 

```{r bac-through-time}
seot <- seot %>% 
  mutate(YEAR = as.numeric(YEAR))

seot %>% 
  filter(FINAL_AGE > 4) %>% 
  ggplot(mapping = aes(x = YEAR, y = BACULA_LGTH, color = BACULA_LGTH)) + 
  geom_jitter() + 
  scale_color_viridis_c() +
  geom_smooth(color = "black") + 
  labs(title = "Baculum Length of Adult Otters", 
       subtitle = "Through the Years", 
       x= "Year", 
       y = "Baculum Length (mm)", 
       color = "Baculum Length")
```

















### Otter's Final Age before and after the Exxon Oil Spill
#### Creating Plots

```{r glimpsing-finalage-otters}
#plot of age at capture before and after oil spill
seot_age_oilspill_graph <- seot %>% 
  filter(oil_spill == "before_oil_spill"|oil_spill == "after_oil_spill") %>% 
  ggplot(mapping = aes(y = FINAL_AGE, x = oil_spill, fill = oil_spill)) + 
  geom_boxplot() +
  scale_fill_manual(values = c("#a1d99b", "#238b45")) +
  labs(title = "Sea Otter Final Age", 
       subtitle = "before and after Exxon Oil Spill", 
       x = "")

seot_age_oilspill_graph
```

#### Check the R-Squared

```{r r-squared-age-oil}
# #now, let's create the linear model comparing age to year
# seot_model_oilage <- linear_reg() %>%
#   set_engine("lm") %>%
#   fit(FINAL_AGE ~ oil_spill, data = seot)
# 
# #let's tidy this model
# seot_model_oilage %>% tidy()
# 
# #augmenting the weight data
# seot_model_oilage_augment <- augment(seot_model_oilage$fit)
# 
# ggplot(seot_model_oilage_augment, mapping = aes(x = .fitted, y = .resid)) +
#   geom_jitter(alpha = 0.75) +
#   geom_smooth(color = "black") +
#   labs(x = "Predicted weight", y = "Residuals")
# 
# #assessing the r-squared value
# glance(seot_model_oilage)$adj.r.squared
```
The r-squared value is `0.04269101`, which means that the variable `oil_spill` accounts for 4.26% of the variation in the dependent variable, `WEIGHT`. 

#### Statistical Analysis

We are performing a statistical analysis on the ages of captured adult otters before and after the Exxon Oil Spill. This is an analysis of two variables (one qualitative, one quantitative), where there are two groups being compared against parametric, independent samples. This suggests that we should use a Welch Two Sample t-test for statistical analysis.

```{r stats-age-oil}
#Welch's t-test
t.test(data = seot, FINAL_AGE ~ oil_spill)
```

REPORT: The otters captured before the oil spill were significantly older than the otters captured after the oil spill (Welch Two Sample t-test; t = -12.825, df = 3472.6, p-value < 0.005).

# Conclusion
The study investigated the effects of the 1989 Exxon Valdez Oil spill on the sea otter population, utilizing data spanning several decades. Findings reveal significant differences in various characteristics of otters captured before and after the spill. Specifically, adult sea otters in the Western Aleutians were notably heavier before the spill, indicating a potential impact on their overall health and condition. Moreover, there was a significant increase in the baculum length of adult male otters post-spill, suggesting possible adaptations or responses to environmental changes. However, body length remained unaffected by the spill. Additionally, otters captured before the spill were notably older than those captured after, indicating potential shifts in population demographics.

