---
title: "Webscraping Attributes"
author: "Autumn Pauly,"
date: "February 2023"
output: github_document
---

#Install/Load Packages

```{r load-packages}
install.packages("robotstxt")
install.packages("rvest")
library(tidyverse)
library(rvest)
library(robotstxt)
library(xml2)
```

## Path Allowed?

We should first determine whether we are allowed to scrape data from this website. 

```{r path-allowed}
paths_allowed("https://alaska.usgs.gov/data/metadata/ecosystems/seaOtter/seot_morphometricsReproStatus_ak_monson/seot_morphometricsReproStatus_ak_monson_metadata.html")
```

### Reading the Page

This will allow us to read the whole page. 

```{r}
attribute_page <- read_html("https://alaska.usgs.gov/data/metadata/ecosystems/seaOtter/seot_morphometricsReproStatus_ak_monson/seot_morphometricsReproStatus_ak_monson_metadata.html#5")

attribute_page
```
### Determining Type of Page
After reading the page, we should see what type of webpage we're looking at.

```{r}
typeof(page)
class(page)
```

# Scraping Website Data

### Scraping Attributes
The function below will allow us to scrape the attribute names. As we can see, these attributes need some string tidying. 

```{r attribute-data}
#scraping attribute title
attribute <- attribute_page |>
  html_nodes(".attrlabl") |>
  html_text()
attribute

#scraping attribute title
attribute_description <- attribute_page |>
  html_nodes(".attrdef") |>
  html_text()
attribute_description
```

Now, let's create the dataframe that we need to display these attributes. 

```{r creating-dataframe}
seot_data_attributes <- tibble(
  attribute = attribute, 
  attribute_description = attribute_description)

seot_data_attributes
```

### Tidying the Data Frame
As we can see, every other row has the information we want. Let's fix this. 

```{r tidying-attribute-dataframe}
seot_data_attributes <- seot_data_attributes %>% 
  filter(attribute != "Attribute_Label:") %>% 
  filter(attribute_description != "Attribute_Definition:") %>% 
  select(attribute, attribute_description)

glimpse(seot_data_attributes)
```

There is still some issues within the `attribute_description` column, where there is a "\n" at the beginning of each value. Let's remove this. 

```{r str-remove}
#removing quotes from column `attribute`
seot_data_attributes <- seot_data_attributes %>% 
  mutate(attribute = str_replace(attribute, '""', '')) %>% 
  mutate(attribute_description = str_replace(attribute_description, '\n', '')) %>% 
  mutate(attribute_description = str_replace(attribute_description, '\n', ''))

seot_data_attributes
```
### Creating Sentence

Now let's create the sentence that we'd need for the ReadME.md file in the `data` folder. 

```{r sentence}
seot_data_attributes <- seot_data_attributes %>%  
  mutate(sentence_description = str_glue("`{attribute}`: {attribute_description}"))
```

Now we have the data/sentences we need to complete the ReadME.md file!